### matricies resumed 
```{r}
set.seed(65)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m
```

```{r}
m[3:4, c(3,1)]
```

### list revisited 
```{r}
xlist <- list(a="BIOL48006220", b=seq(1,10, by = .5), data = "grade")
xlist
```


```{r}
xlist[1:2]
```


```{r}
xlist['a']
```


```{r}
xlist['b']
```


```{r}
xlist[['b']]
```

```{r}
xlist$b
```


### data frames revisited 
```{r}
gp <- read.csv('/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/data /sdi.csv')
```

```{r}
head(gp[3], n=10L)
```
```{r}
nrow(gp)
```
```{r}
head(gp[["country"]],10L)
```

```{r}
gp$X2016
```


```{r}
gp[c(1,3), 2:5]
```

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2018 <= 100),]
```


### conditionals 
```{r}
n <- 10

if (n<10) {
  print("n is less than 10")
  } else if (n>10) {
    print("n is greater than 10")
  } else { 
    print("n is equal to 10")}
  
```

```{r}
for (n in seq(1,20)){
  

  if (n<10) {
    print("n is less than 10")
    } else if (n>10) {
     print("n is greater than 10")
    } else { 
      print("n is equal to 10")}}
```
```{r}
g <- 0

while(g<=10) { 
  print(paste(g, "is less than or equal to 10"))
  g <- g+1
  }
```


## plotting and figures 
```{r}
# install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

```{r}
gp <- read.csv('/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/data/gapminder_all.csv')

head(gp)
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")

head(gapminder)

ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()

ggplot(data=gapminder, mapping=aes(x=year, lifeExp, by=country, color=continent)) + geom_line()

```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, lifeExp, by=country, color=continent)) + geom_line()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping=aes(color=continent)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25) + scale_x_log10()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.3, color="purple", size=1.3) + scale_x_log10() + geom_smooth(method=lm, color="gold")

```


```{r}
africas <- gapminder[gapminder$continent == "Africa", ]
head(africas)
```


```{r}
ggplot(data = africas, mapping = aes(x=year, y=lifeExp)) + 
  geom_line() +
  facet_wrap( ~ country) + 
  theme(axis.text.x = element_text(angle = 45)) + 
  labs(
    x = "Year",
    y = "Life Expectency", 
    title = "Life Expectency Over Time in African Countries"
  )
```


```{r}
AfricanLifeExpectency <- ggplot(data = africas, mapping = aes(x=year, y=lifeExp)) + 
  geom_line() +
  facet_wrap( ~ country) + 
  theme(axis.text.x = element_text(angle = 45)) + 
  labs(
    x = "Year",
    y = "Life Expectency", 
    title = "Life Expectency Over Time in African Countries"
  )
```


```{r}
AfricanLifeExpectency
ggsave(filename = "/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/results/AfricanLifeExp.png", plot = AfricanLifeExpectency, width = 24, height = 40, dpi = 300, units = "cm")
```

```{r}
pdf(file = "/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/results/AfricanLifeExp.pdf", width = 24, height = 40)
plot(AfricanLifeExpectency)
dev.off()

```


```{r}
write.table(gapminder, file = "/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/data/gapminder_web", sep = ",")
```

```{r}
write.csv(africas, file = "/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/data/gapminder_web_africas")
```


```{r}
### install.packages(c("ggridges", "viridis", "hrbrthemes"), dependencies = T)
```

```{r}
library(ggridges)
library(viridis)
library(hrbrthemes)
library(ggplot2)
```

```{r}
## install.packages(c("ggstatsplot", "palmerpenguins", "tidyverse"), dependencies = T)
```

```{r}
library(ggstatsplot)
library(palmerpenguins)
library(tidyverse)
```

```{r}
penguins
```


```{r}
gapminder2 <- drop_na(gapminder)
```



### 10/27 class - statistics in R

```{r}
median(gapminder$lifeExp)
```

```{r}
## variation 
    # range 
range()

    # difference 
diff()

    # variance 
var()

    # std dev 
sd()

    #IQR
IDR()

    # z - score 
scale() 

    # summary 
summary() 

```



# basic R plots 

```{r}
# histogram 
hist() 

#plots 
plot()

# box plots 
boxplot()
```

# bi/multivariate stats 

```{r}
# plotting and regression 
 
spid.gen <- read.csv("/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/data/spider_genitalia.csv", header = T)
```


```{r}
# box plots 
boxplot(spid.gen$left.bulb ~ spid.gen$habitat)
```




```{r}
# scatter plots 
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch = as.character(spid.gen$habitat))
```

```{r}
# linear regression 
gen.reg <- lm(spid.gen$left.bulb ~ spid.gen$right.bulb)
print(gen.reg)
summary(gen.reg)
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
abline(gen.reg)
```

```{r}
# correlation coefficients & spearman rank correlation 
cor.gen <- with(spid.gen, cor(left.bulb, right.bulb))
print(cor.gen^2)
```

```{r}
spear.cor.gen <- with(spid.gen, cor(left.bulb, right.bulb, method="spearman"))
print(spear.cor.gen)
```

```{r}
# residuals 
resid(gen.reg)
```
```{r}
# transformations 
plot(spid.gen$left.bulb^2~spid.gen$right.bulb^2)
```

2. comparing discrete treatment effects 

```{r}
# chi square 
exp <- c(.16,.17,.16,.17,.17,.17)
obs_weighted <- c(4,15,6,15,18,2)
obs_fair <- c(10,10,10,10,10,10)
chisq.test(obs, p=exp)
chisq.test(obs_fair, p=exp)
```


```{r}
# t tests 
#H0 - true mean is equal to 0, but you can set true mean equal to other values with mu=
t.test(spid.gen$carapace.length)
mean(spid.gen$carapace.length)
t.test(spid.gen$carapace.length, mu=29)
```
```{r}
# two sample t test 
habitat.t2 <- t.test(spid.gen$carapace.length ~ spid.gen$habitat)
habitat.t2
```

```{r}
habitat.tless <- t.test(spid.gen$left.bulb, alternative = "less")
habitat.tless
habitat.tgreater <- t.test(spid.gen$left.bulb, alternative = "greater")
habitat.tgreater

```



# plotting 

```{r}
plot(1:25, rep(.25,25), pch=1:25, col=1:25, ylim=c(0,6), cex=2, ylab="line types (lty) 1 to 6", xlab="plotting character = (pch) 1 to 25 and colours (col) 1 to 8", main= "line types (lty), plotting characters (pch), /nand colors (col) for plot and xyplot", lab=c(25,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=3)

abline(h=1:6, lty=1:6, col=1:6, lwd=3)
```

```{r}
pdf("/Users/bolajohn/Desktop/ECU/biol48006220_fall2022/practcomp_2022/results/plotCheatSheet.pdf")

plot(1:25, rep(.25,25), pch=1:25, col=1:25, ylim=c(0,6), cex=2, ylab="line types (lty) 1 to 6", xlab="plotting character = (pch) 1 to 25 and colours (col) 1 to 8", main= "line types (lty), plotting characters (pch), /nand colors (col) for plot and xyplot", lab=c(25,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=3)

abline(h=1:6, lty=1:6, col=1:6, lwd=3)

dev.off()
```

## looping and conditionals 
a. if then statements 
```{r}
x <- 0
if(x>0) {
  print("positive number")
} else if (x<0) {
  print("negative number")
} else 
  print("zero")
```

```{r}
a <- c(5,7,2,9)
ifelse(a %% 2 == 0, "even", "odd")
```

b. for loops 
```{r}
z <- c(2,5,3,9,8,11,6)
count <- 0 
for (val in z) { 
  if(val %% 2 == 0) count = count+1
}

print(count)


```

c. While statement 
```{r}
i <- 1 
while (i<5) {
  print(i)
  i = i+1
}
```


d. interruptions
```{r}
x <- 1:5 

for (val in x) {
  if (val == 3) { 
    break
  }
  print(val)
}

```

```{r}
x <- 1:5
for (val in x) {
  if (val == 3) {
    next
  }
  print(val)
}
```

e. repeat 
```{r}
x <- 1 
repeat { 
  print(x)
  x = x+1
  if (x == 20) {
    break
  }
   
  }
```


## randomization and dataset management (short)
a. data management 
```{r}
B <- matrix( 
  c(2,4,3,1,5,87),
  nrow=3,
  ncol=2) 

B

```


```{r}
# transpose matrices 
t(B)
```

```{r}
C <- matrix( 
  c(7,4,2),
  nrow=3,
  ncol=1
  )

C
```


```{r}
## combining data 
BC <- cbind(B, C)
```

```{r}
## deconstruct to vextor of values 

c(B)
```


b. randomizations
```{r}
# r norm(n, mean, sd)
# sample(x, size, replace = F/T, prob = null)

ndist <- rnorm(70, 10, 10)
ndist
```


```{r}
mean(ndist)
sd(ndist)
```

```{r}
hist(ndist)
```

```{r}
sample(ndist,10, replace = F)
```


```{r}
## random number generator for subsampling 
test <- 1:70
sample(test, 100, replace = T)
```

```{r}
sample(test, 70, replace = T)
```

```{r}
test1 <- c(391,392,393,394,395,396,19)
sample(test1, 5, replace = T)
```



